Kindly Bent To Free Us

### 2020-08-06 kindly/slides-icfp.tex
Apply peter's remarks

\begin{frame}
  \frametitle{Motivation}

  Memory safety is the cause of numerous bugs and security issues.

  Recent classification (2015-2020) of\\ \og high severity security bugs\fg in Chromium:

  \begin{figure}[h]
    \centering
    \includegraphics[width=0.5\textwidth]
    {chromium-use-after-free}
  \end{figure}

  \pause
  Chromium is written in C/C++!\\
  Surely these bugs don't happen in high-level typed languages\alt<3>{ \dots right?}{.}

@@

\begin{minted}{OCaml}
let gradeDB : database = Dbm.opendbm "gradeDB" ... in
Dbm.add gradeDB "math" 42;
(* ... *)
Dbm.close gradeDB;
(* ... *)
print_int (Dbm.find gradeDB "literature") (* bug! *)
\end{minted}

\begin{frame}
  \frametitle{Solutions?}

  \TODO{Existing ways to solve this}

@@

  \begin{onlyenv}<1-3>
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let gradeDB : #\tikzcoord{db1}#database#\tikzcoord{db2}# = Dbm.opendbm "gradeDB" ... in
Dbm.add #\tikzcoord{A1}#&!gradeDB#\tikzcoord{A2}# #\tikzcoord{string1}#"math"#\tikzcoord{string2}# 42;
(* ... *)
Dbm.close gradeDB;
(* ... *)
print_int (Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "literature") (* \N Compile time error! *)
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<4>
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let gradeDB #\tikzcoord{eq1}#=#\tikzcoord{eq2}# Dbm.opendbm "gradeDB" ... in
Dbm.add &!gradeDB "math" 42;
(* ... *)
Dbm.close gradeDB;
(* ... *)
print_string (Dbm.find &gradeDB "literature") (* \N Compile time error! *)
\end{minted}
  \end{onlyenv}

@@

  \textbf{Examples:}
  \begin{onlyenv}<1>
\begin{minted}[texcomments=true]{OCaml}
type database : lin
type string : un
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<2>
\begin{minted}[texcomments=true]{OCaml}
type ('a : 'k) list : 'k
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<3>
\begin{minted}[texcomments=true]{OCaml}
type ('a : 'k) list : 'k
val create_list : ('a : un) => int -> 'a -> 'a list
\end{minted}

    We also use \textbf{kind constraints} and \textbf{subkinding}\\
    $\Rightarrow$ Type signatures are still reasonably simple!
    \TODO{More on subkinding ?}
  \end{onlyenv}
\end{frame}

@@

  What about functions ? We can capture: 
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let gradeDB = Dbm.open ...

let #\tikzcoord{fun1}#log_n_close#\tikzcoord{fun2}# msg = 
  printf "Closing: %s" msg;
  Dbm.close gradeDB
\end{minted}
  \begin{onlyenv}<3->
    We infer the type:
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
val conclude : string #\tikzcoord{arr1}##$\tarr{\texttt{lin}}$##\tikzcoord{arr2}# unit
\end{minted}
  \end{onlyenv}

  \begin{tikzpicture}[remember picture,overlay]
    \begin{onlyenv}<2>
      \node[code,fit=(fun1) (fun2)] (fun) {};
      \draw[overlay,link,<-] (fun) to[out=30,in=180]
      +(3,0.2) node[annot,scale=1,anchor=west,text width=3.2cm] {
        This function \textbf{must}\\ be used exactly once
      };
    \end{onlyenv}
    \begin{onlyenv}<3->
      \node[code,minimum height=18pt,fit=(arr1) (arr2)] (arr) {};
      \draw[overlay,link,<-] (arr) to[out=-110,in=70]
      +(-1,-1) node[annot,scale=1,anchor=north] {
        Arrows are annotated with their usage mode
      };
      \node<4>[annot,scale=1.2,red] at ($(arr)+(4,1)$) {
        \textbf{Warning}: This does not say anything about the arguments!!
      };
    \end{onlyenv}
  \end{tikzpicture}
\end{frame}

@@

  Borrows allow to lend usage of something to someone else.
  \begin{itemize}
  \item \textbf{Shared} borrows \tikzcoord{UX1}@&a@\tikzcoord{UX2} which are Unrestricted (@un@)
  \item \textbf{Exclusive} borrows \tikzcoord{AX1}@&!a@\tikzcoord{AX2} which are Affine (@aff@)
  \end{itemize}

  \begin{onlyenv}<2->
  A correct usage of borrows:
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let avg =
  (Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "math" + Dbm.find #\tikzcoord{U3}#&gradeDB#\tikzcoord{U4}# "compsci") / 2
  (* \Y Multiple shared borrows *)
in
Dbm.add #\tikzcoord{A1}#&!gradeDB#\tikzcoord{A2}# "average" avg (* \Y One exclusive borrow *)
\end{minted}
  \end{onlyenv}
  \visible<3>{There are rules to respect!}

  \begin{tikzpicture}[remember picture,overlay]
    \begin{onlyenv}<2->
      \node[code,color=yellow,fit=(A1) (A2)] {};
      \node[code,color=red,fit=(U1) (U2)] {};
      \node[code,color=red,fit=(U3) (U4)] {};
    \end{onlyenv}
    \node[code,color=yellow,fit=(AX1) (AX2)] {};
    \node[code,color=red,fit=(UX1) (UX2)] {};
  \end{tikzpicture}
\end{frame}

@@

% \begin{onlyenv}<2>
% \begin{minted}[texcomments=true]{OCaml}
% let gradeDB = ... in
% let avg =
%  let grade topic = Dbm.find &gradeDB topic in (* \Y Capture *)
%  (grade "math" + grade "compsci") / 2
% in
% Dbm.add &!gradeDB "average" avg (* \Y Exclusive borrow *)
% \end{minted}
% \end{onlyenv}
\begin{onlyenv}<1>
  \textbf{Rule 1}: Cannot use a borrow and the thing itself simultaneously
\begin{minted}[texcomments=true]{OCaml}
let gradeDB = ... in
f (gradeDB, &gradeDB) (* \N Conflicting use and borrow! *)
\end{minted}
\end{onlyenv}%
\begin{onlyenv}<2>
  \textbf{Rule 2}: Cannot use an exclusive borrow and any other borrow simultaneously
\begin{minted}[texcomments=true]{OCaml}
let gradeDB = ... in
f (&!gradeDB, &gradeDB) (* \N Conflicting borrows! *)
\end{minted}
\end{onlyenv}%
\begin{onlyenv}<3->
  \textbf{Rule 3}: Borrows must not escape
  \begin{onlyenv}<3>
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let f () = 
  let gradeDB = ... in
  #\tikzcoord{R1}#let x = (&gradeDb, "mygrades") in#\tikzcoord{R2}#
  #\tikzcoord{R3}#x#\tikzcoord{R4}#
  (* \N Borrow escaping its scope! *)
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<4->
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let f () = 
  let gradeDB = ... in
  #\tikzcoord{R1}#{| let x = (#\tikzcoord{B1}#&gradeDb#\tikzcoord{B2}#, "mygrades") in#\tikzcoord{R2}#
  #\tikzcoord{R3}#x |}#\tikzcoord{R4}#
  (* \N Borrow escaping its scope! *)
\end{minted}
  \end{onlyenv}%
  \begin{tikzpicture}[remember picture,overlay]%
    \begin{onlyenv}<4->
      \node[code,rounded corners=0,color=red,fit=(R1) (R2)] {};
      \node[code,rounded corners=0,minimum height=15pt,yshift=1pt,color=red,fit=(R3) (R4)] {};
      \node[code,opacity=0,fit=(B1) (B2)] (B) {};
    \end{onlyenv}
    \begin{onlyenv}<5->
      \draw[link,<-] (B.10) to[in=180,out=40] ++(0.5,0.5) node[annot, anchor=west] {
        @&gradeDb : &(database, un_2)@
      };
      \draw[link,<-] (R2) to[in=90,out=-10] ++(1.5,-0.5) node[annot, anchor=north] {
        Region nesting level: 1
      };
    \end{onlyenv}
  \end{tikzpicture}%
  \visible<5>{%
    We use \textbf{indexed} kinds to check that borrows do not escape!

@@

  Inference proceeds in the following steps:

  \begin{enumerate}
  \item Infer where regions goes
  \item Turn the type, linearity and escape checks into constraints
  \item Solve the constraints
  \item Simplify the resulting types
  \end{enumerate}
\end{frame}

@@

      \draw[link,->] (S1) to node[midway,above,sloped] {1. Infer regions} (S2) ;

      \visible<2->{
        \node[annot,text width=4.8cm] at (8,0) (S3) {%
          \vspace{-1em}
          \begin{align*}
            \E = &(\tvar_f : \kvar_f)
                 (\tvar_x : \kvar_x)\dots\\
            C = &(\tvar_f \leq \gamma \tarr{\kvar_1} \beta )
                \Cand
                (\gamma \leq \tvar_x)\\
                &\Cand
                (\beta \times  \tvar_x \leq \alpha_r)
                \Cand
                  (\kvar_x \leq \kun)\\
                 &\Cand \dots
          \end{align*}
        };
        \draw[link,->] (S2) to
        node[midway,right,text width=3cm] {2. Convert\\ into constraints}
        (S3);
      }

@@

  \only<1>{%
    1. Infer the placement of region based on the position of borrows and the borrowing rules.
  }%
  \only<2>{%
    2. Convert into a custom constraint system based on HM(X).
  }%
  \only<3>{%
    3. Solve the constraints using a custom algorithm and obtain principal type schemes.
  }%
  \only<4>{%
    4. Simplify the obtained type scheme by leveraging subkinding
  }%

@@

\begin{frame}
  \frametitle{The theory -- Summary}
  In the paper, the theoretical contributions:

>>>>

\begin{frame}
  \frametitle{\og high severity security bugs\fg in Chromium}

  \begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]
    {chromium-use-after-free}
  \end{figure}

  \visible<2->{
    Chromium is written in C/C++!\\
    Surely these bugs don't happen in high-level typed languages%
    \alt<3>{ \dots right?}{.}
  }

@@

\begin{minted}{OCaml}
let gradeDB : database = Dbm.opendbm "gradeDB" ... in
Dbm.add gradeDB "math" 42;
(* ... *)
Dbm.close gradeDB;
(* ... *)
print_int (Dbm.find gradeDB "literature") (* run-time error! *)
\end{minted}

@@

  \begin{onlyenv}<1-3>
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let gradeDB : #\tikzcoord{db1}#database#\tikzcoord{db2}# = Dbm.opendbm "gradeDB" ... in
Dbm.add #\tikzcoord{A1}#&!gradeDB#\tikzcoord{A2}# #\tikzcoord{string1}#"math"#\tikzcoord{string2}# 42;
(* ... *)
Dbm.close gradeDB;
(* ... *)
print_int (Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "literature") (* \N compile-time error! *)
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<4>
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let gradeDB #\tikzcoord{eq1}#=#\tikzcoord{eq2}# Dbm.opendbm "gradeDB" ... in
Dbm.add &!gradeDB "math" 42;
(* ... *)
Dbm.close gradeDB;
(* ... *)
print_string (Dbm.find &gradeDB "literature") (* \N compile-time error! *)
\end{minted}
  \end{onlyenv}

@@

  \textbf{Examples:}
  \begin{onlyenv}<1>
\begin{minted}[texcomments=true]{OCaml}
type database : lin
type string : un
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<2>
\begin{minted}[texcomments=true]{OCaml}
type ('a : 'k) list : 'k
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<3->
\begin{minted}[texcomments=true]{OCaml}
type ('a : 'k) list : 'k
val create_list : ('a : un) => int -> 'a -> 'a list
\end{minted}

    \begin{onlyenv}<4>
      We also use \textbf{subkinding}: @un <= aff <= lin@
    \end{onlyenv}
  \end{onlyenv}
\end{frame}

@@

\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let gradeDB = Dbm.open ...

let #\tikzcoord{fun1}#log_n_close#\tikzcoord{fun2}# msg = 
  printf "Closing: %s" msg;
  Dbm.close #\tikzcoord{db1}#gradeDB#\tikzcoord{db2}#
\end{minted}
  \begin{onlyenv}<3->
    We infer the type:
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
val log_n_close : string #\tikzcoord{arr1}##$\tarr{\texttt{lin}}$##\tikzcoord{arr2}# unit
\end{minted}
  \end{onlyenv}

  \begin{tikzpicture}[remember picture,overlay]
    \begin{onlyenv}<2->
      \node[code,fit=(fun1) (fun2)] (fun) {};
      \node[code,fit=(db1) (db2)] (db) {};
      \node[annot,scale=1.2]
      at ($(fun)+(5,0.2)$) (capture) {
        Capture!
      };
      \draw[overlay,link] (fun) to[out=20,in=170] (capture) ;
      \draw[overlay,link] (capture) to[out=-90,in=0] (db) ;
    \end{onlyenv}
    \begin{onlyenv}<3->
      \node[code,minimum height=18pt,fit=(arr1) (arr2)] (arr) {};
      \draw[overlay,link,<-] (arr) to[out=-110,in=70]
      +(-1,-1) node[annot,scale=1.2,anchor=north] {
        Usage mode
      };
    \end{onlyenv}
    \begin{onlyenv}<4>
      \node[annot,scale=1.2,red] at ($(arr)+(4,1)$) (warn) {
        \textbf{Warning}: Does not say anything about the arguments!!
      };
      \draw[overlay,link,<-,red] (arr.north) to[out=70,in=-150] (warn.-175) ;
    \end{onlyenv}
  \end{tikzpicture}
\end{frame}

@@

  A borrow is a temporary loan of a resource @a@
  \begin{itemize}
  \item
    \textbf{Shared} borrows \tikzcoord{UX1}@&a@\tikzcoord{UX2} are for observing the resource (Unrestricted -- @un@)
    % \textbf{Shared} borrows \tikzcoord{UX1}@&a@\tikzcoord{UX2} which are Unrestricted (@un@)
  \item
    \textbf{Exclusive} borrows \tikzcoord{AX1}@&!a@\tikzcoord{AX2} are for modifying the resource (Affine -- @aff@)
    % \textbf{Exclusive} borrows \tikzcoord{AX1}@&!a@\tikzcoord{AX2} which are Affine (@aff@)
  \end{itemize}

  \begin{onlyenv}<2->
  A correct usage of borrows:
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let avg =
  (Dbm.find #\tikzcoord{U1}#&gradeDB#\tikzcoord{U2}# "math" + Dbm.find #\tikzcoord{U3}#&gradeDB#\tikzcoord{U4}# "compsci") / 2
  (* \Y Multiple shared borrows *)
in
Dbm.add #\tikzcoord{A1}#&!gradeDB#\tikzcoord{A2}# "average" avg (* \Y One exclusive borrow *)
\end{minted}
  \end{onlyenv}  
  \begin{tikzpicture}[remember picture,overlay]
    \begin{onlyenv}<2->
      \node[code,color=yellow,fit=(A1) (A2)] {};
      \node[code,color=red,fit=(U1) (U2)] {};
      \node[code,color=red,fit=(U3) (U4)] {};
    \end{onlyenv}
    \node[code,color=yellow,fit=(AX1) (AX2)] {};
    \node[code,color=red,fit=(UX1) (UX2)] {};
  \end{tikzpicture}
\end{frame}

@@

% \begin{onlyenv}<2>
% \begin{minted}[texcomments=true]{OCaml}
% let gradeDB = ... in
% let avg =
%  let grade topic = Dbm.find &gradeDB topic in (* \Y Capture *)
%  (grade "math" + grade "compsci") / 2
% in
% Dbm.add &!gradeDB "average" avg (* \Y Exclusive borrow *)
% \end{minted}
% \end{onlyenv}
\begin{onlyenv}<1>
  \textbf{Rule 1}: Cannot use a borrow and the resource itself simultaneously
\begin{minted}[texcomments=true]{OCaml}
let gradeDB = ... in
f (gradeDB, &gradeDB) (* \N Conflicting use and borrow! *)
\end{minted}
\end{onlyenv}%
\begin{onlyenv}<2>
  \textbf{Rule 2}: Cannot use an exclusive borrow and any other borrow simultaneously
\begin{minted}[texcomments=true]{OCaml}
let gradeDB = ... in
f (&!gradeDB, &gradeDB) (* \N Conflicting borrows! *)
\end{minted}
\end{onlyenv}%
\begin{onlyenv}<3->
  \textbf{Rule 3}: Borrows must not escape
  \begin{onlyenv}<3>
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let f () = 
  let gradeDB = ... in
  #\tikzcoord{R1}#let x = (&gradeDb, "mygrades") in#\tikzcoord{R2}#
  #\tikzcoord{R3}#x#\tikzcoord{R4}#
  (* \N Borrow escaping its scope! *)
\end{minted}
  \end{onlyenv}
  \begin{onlyenv}<4->
\begin{minted}[texcomments=true,escapeinside=\#\#]{OCaml}
let f () = 
  let gradeDB = ... in
  #\tikzcoord{R1}#{| let x = (#\tikzcoord{B1}#&gradeDb#\tikzcoord{B2}#, "mygrades") in#\tikzcoord{R2}#
  #\tikzcoord{R3}#x |}#\tikzcoord{R4}#
  (* \N Borrow escaping its scope! *)
\end{minted}
  \end{onlyenv}%
  \begin{tikzpicture}[remember picture,overlay]%
    \begin{onlyenv}<4->
      \node[code,rounded corners=0,color=red,fit=(R1) (R2)] {};
      \node[code,rounded corners=0,minimum height=15pt,yshift=1pt,color=red,fit=(R3) (R4)] {};
      \node[code,opacity=0,fit=(B1) (B2)] (B) {};
    \end{onlyenv}
    \begin{onlyenv}<5->
      \draw[link,<-] (B.10) to[in=180,out=40] ++(0.5,0.5) node[annot, anchor=west] {
        @&gradeDb : &(database, un_2)@
      };
      \draw[link,<-] (R2) to[in=90,out=-10] ++(1.5,-0.5) node[annot, anchor=north] {
        Region nesting level: 1
      };
    \end{onlyenv}
  \end{tikzpicture}%
  \visible<5>{%
    \textbf{Indexed} kinds ensure that borrows do not escape!

@@

  \begin{enumerate}
  \item Elaborate regions
  \item Generate constraints for type, linearity and escape checks
  \item Solve the constraints
  \item Simplify the resulting types
  \end{enumerate}
\end{frame}

@@

      \draw[link,->] (S1) to node[midway,above,sloped] {1. Elaborate regions} (S2) ;

      \visible<2->{
        \node[annot,text width=4.8cm] at (8,0) (S3) {%
          \vspace{-1em}
          \begin{align*}
            \E = &(\tvar_f : \kvar_f)
                 (\tvar_x : \kvar_x)\dots\\
            C = &(\tvar_f \leq \gamma \tarr{\kvar_1} \beta )
                \Cand
                (\gamma \leq \tvar_x)\\
                &\Cand
                (\beta \times  \tvar_x \leq \alpha_r)
                \Cand
                  (\kvar_x \leq \kun)\\
                 &\Cand \dots
          \end{align*}
        };
        \draw[link,->] (S2) to
        node[midway,right,text width=3cm] {2. Generate\\ constraints}
        (S3);
      }

@@

  \only<1>{%
    1. Infer the placement of region based on the position of borrows and the borrowing rules.
  }%
  \only<2>{%
    2. Generate custom constraints based on HM(X).
  }%
  \only<3>{%
    3. Solve the constraints using a custom algorithm and obtain principal type schemes.
  }%
  \only<4>{%
    4. Simplify the obtained type scheme by leveraging subkinding
  }%

@@

\begin{frame}
  \frametitle{The theory -- Summary}
